---
title: "ProjectAssignment2"
author: "yangru1q1"
date: '2019-07-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
library(leaflet)
```

## Synopsis

It is hot in summer, especially for someone who does not have an air-conditioning. In this project, I'm using [Air Conditioned Public Places & Cooling Centres](https://www.toronto.ca/city-government/data-research-maps/open-data/open-data-catalogue/locations-and-mapping/#d7f09371-1386-cc5c-69c9-79d4cf7f6faa) dataset to show the availability of buildings that are open to the general public that offer an air-conditioned space for temporary relief from the extreme heat in **Toronto** area.

## Map
```{r, echo=F}
url <- "http://app.toronto.ca/opendata//ac_locations/locations.csv?v=1.00"

ac <- read.csv(url)

ac.data <- ac[, -c(1, 2, 3)]

ac.data$pop <- with(ac.data, paste("Location Name:", Location.Name, "<br>",
                                   "Type:", Location.desc, "<br>",
                                   "Adress:", Address, "<br>",
                                   "Phone:", Phone))


url1 <- "https://emojipedia-us.s3.dualstack.us-west-1.amazonaws.com/thumbs/320/apple/198/freezing-face_1f976.png"
url2 <- "https://emojipedia-us.s3.dualstack.us-west-1.amazonaws.com/thumbs/320/google/146/freezing-face_1f976.png"
url3 <- "https://emojipedia-us.s3.dualstack.us-west-1.amazonaws.com/thumbs/320/microsoft/209/freezing-face_1f976.png"
url4 <- "https://emojipedia-us.s3.dualstack.us-west-1.amazonaws.com/thumbs/320/samsung/161/freezing-face_1f976.png"


coldIcon <- icons(
        iconUrl = ifelse(ac.data$Location.desc == "Library", url1,
                         ifelse(ac.data$Location.desc == "Community Centre", url2,
                                ifelse(ac.data$Location.desc == "Civic Centre", url3, url4))),
        iconWidth = 30, iconHeight = 30)


ac.data %>% leaflet() %>%
        addTiles() %>%
        addMarkers(lat = ~Lat, 
                   lng = ~Lon,
                   popup = ~pop,
                   icon = coldIcon)
```

## Conclusion
Let's all remember [Willis Carrier](https://en.wikipedia.org/wiki/Willis_Carrier), this handsome guy saved our life.

